---
title: "Attrition and Salary Analysis - Case Study 2"
author: "Roslyn Smith"
date: "4/16/2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


Importing packages 
```{r,message = FALSE,warning = FALSE, echo=FALSE}
#Libraries
library(tidyverse)
library(caret)
library(class)
library(e1071)
library(maps)
library(mapproj)
library(plotly)
library(data.table)
library(formattable)
library(tidyr)
library(dplyr)
library(GGally)
library(ggplot2)
library(magrittr)
library(scatterplot3d)
library(leaps)
library(PerformanceAnalytics)
library(olsrr)
library(dplyr)
library(FNN) 
library(gmodels) 
library(psych)
library(Metrics)

```

Importing datasets
```{r,message = FALSE,warning = FALSE, echo=FALSE}
CS2_data = read.csv("C:/Users/rosjo/Documents/Masters Program/Doing Data Science/Assignments/Case Study 2/CaseStudy2-data.csv",header = TRUE, stringsAsFactors = TRUE)
var.names.CS2_data <-tolower(colnames(CS2_data))
colnames(CS2_data) <- var.names.CS2_data


CS2_attrition_test = read.csv("C:/Users/rosjo/Documents/Masters Program/Doing Data Science/Assignments/Case Study 2/CaseStudy2CompSet No Attrition.csv",header = TRUE, stringsAsFactors = TRUE)
var.names.CS2_attrition_test <-tolower(colnames(CS2_attrition_test))
colnames(CS2_attrition_test) <- var.names.CS2_attrition_test



CS2_salary_test = read.csv("C:/Users/rosjo/Documents/Masters Program/Doing Data Science/Assignments/Case Study 2/CaseStudy2CompSet No Salary.csv",header = TRUE, stringsAsFactors = TRUE)
var.names.CS2_salary_test <-tolower(colnames(CS2_salary_test))
colnames(CS2_salary_test) <- var.names.CS2_salary_test

CS2_salary_test$attrition2 <-ifelse(CS2_salary_test$attrition =="Yes",1,0)

```

<font size="6"> Adjustments to data provided </font>

I excluded employee unique identifiers as well as columns with same data from the analysis to better group employees and reduce dataset.  In addition, I converted the Attrition Yes / No indicators to binary 0/1 in order to better run a model.

Columns Excluded:
  Employee Unique Identifiers
    - ID and Employee Number
  Columns with no differentiation in data
    - Employee Count: Value is all 1
    - Over 18: Value is all Yes
    - Standard Hours: Value is all 80 hours

```{r,message = FALSE,warning = FALSE, echo=FALSE}
#create dataframe for model
model_data <- CS2_data

#exclude unique associate ids and columns with same data in each row
model_data <- model_data %>% dplyr::select(-id, -employeecount, -employeenumber, -over18, -standardhours)

## Convert Attrition to numeric for models
model_data$attrition2 <-ifelse(model_data$attrition =="Yes",1,0)
```


<font size="6"> Initial Attrition Model</font>

Create an initial model to determine significant variables that impact attrition
```{r,message = FALSE,warning = FALSE, echo=FALSE}
## Linear model to identify most statistically significant variables

model_attrition <- lm(attrition2~. - attrition,data = model_data)
summary(model_attrition)

### Adjusted R2: 0.25
## Most significant variables: Overtime - Yes (p-value <0.0001), Job Involvement (p-value <0.0001), Job Satisfaction(p-value <0.0001), Num Companies worked (p-value <0.0001) and Marital Status - Single(p-value <0.0008)



```
In running a linear regression model with all variables, the variables with the most significance were:
  - Overtime completed
  - Job Involvement
  - Job Satisfaction
  - Number of Companies Worked
  - Marital Status


<font size="6"> Plotting Attrition vs. Key Variables</font>

```{r,message = FALSE,warning = FALSE, echo=FALSE}
##Plotting Attrition vs. variables

attrition<-factor(model_data$attrition,c("Yes","No"),labels=c('Attrited','Retained')) 
overtime<-factor(model_data$overtime,c("Yes","No"),labels=c('Had Overtime','No Overtime'))
jobinvolvement<-factor(model_data$jobinvolvement,levels=c(1,2,3,4),labels=c('Not Involved','Somewhat Involved','Involved','Very Involved'))  
jobsatisfaction<-factor(model_data$jobsatisfaction,levels=c(1,2,3,4),labels=c('Not Satisfied','Somewhat Satisfied','Satisfied','Very Satisfied'))
numcompaniesworked<-factor(model_data$numcompaniesworked,levels=c(0,1,2,3,4,5,6,7,8,9),labels=c('None','One','Two','Three','Four','Five','Six','Seven','Eight','Nine'))
maritalstatus<-model_data$maritalstatus
businesstravel<-as.factor(model_data$businesstravel)
jobrole<-model_data$jobrole
joblevel<-factor(model_data$joblevel,levels=c(1,2,3,4,5),labels=c('Level 1','Level 2','Level 3','Level 4','Level 5'))


##Plotting Attrition vs. Overtime
cross_ot <- table(attrition, overtime)
plot_data_ot<- as.data.frame(round(prop.table(cross_ot,2)*100, digits = 0))

plot_data_ot%>%
  ggplot(aes(x=overtime, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 2, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Overtime')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))
```

The column chart shows a higher percentage of employees who worked overtime also attrited.


```{r,message = FALSE,warning = FALSE, echo=FALSE}
##Plotting Attrition vs. Job Involvement
cross_ji <- table(attrition, jobinvolvement)
plot_data_ji<- as.data.frame(round(prop.table(cross_ji,2)*100, digits = 0))

plot_data_ji%>%
  ggplot(aes(x=jobinvolvement, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 2, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Job Involvement')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))
```

As job involvement increases, we see a decrease in attrition.  



```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Attrition vs. Job Satisfaction
cross_js <- table(attrition, jobsatisfaction)
plot_data_js<- as.data.frame(round(prop.table(cross_js,2)*100, digits = 0))

plot_data_js%>%
  ggplot(aes(x=jobsatisfaction, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 2, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Job Satisfaction')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))
```

Similar to job involvement, we see an inverse trend with job satisfaction and attrition.



```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Attrition vs. Number of Companies Worked
cross_ncw <- table(attrition, numcompaniesworked)
plot_data_ncw<- as.data.frame(round(prop.table(cross_ncw, margin=2)*100, digits=0))

plot_data_ncw%>%
  ggplot(aes(x=numcompaniesworked, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 1, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Number of Companies Worked')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))
```

As there is an increase in the number of companies the employees worked with, we see an increasing trend of attrition.



```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Attrition vs. Marital Status
cross_ms <- table(attrition, maritalstatus)
plot_data_ms<- as.data.frame(round(prop.table(cross_ms, margin=2)*100, digits=0))

plot_data_ms%>%
  ggplot(aes(x=maritalstatus, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 2, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Marital Status')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+
  scale_x_discrete(limits=c("Single","Married","Divorced"))+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))
```

Employees’ personal life also impact attrition.  There is a higher proportion of single employees that attrited when compared to married or divorced employees.



```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Attrition vs. Job Role
cross_jr <- table(attrition, jobrole)
plot_data_jr<- as.data.frame(round(prop.table(cross_jr, margin=2)*100, digits=0))

plot_data_jr%>%
  ggplot(aes(x=jobrole, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 1, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Job Role')+
  ylab('Percentage')+
  scale_x_discrete(labels = function(jobrole) str_wrap(jobrole, width = 10))+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust =1, size = 10))
```

There is significantly higher attrition in the Sales Representative role which is not replicated in any other role.


```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Attrition vs. Business Travel
cross_bt <- table(attrition, businesstravel)
plot_data_bt<- as.data.frame(round(prop.table(cross_bt, margin=2)*100, digits=0))

plot_data_bt%>%
  ggplot(aes(x=businesstravel, y=Freq, fill = attrition))+ geom_bar(stat="identity", position = "dodge")+
  geom_text(aes(label  = paste0(Freq,"%")), vjust = 2, size = 3.2, color = "black", position = position_dodge(width = 0.9))+
  ggtitle('Percentage Attrition by Business Travel Frequency')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Attrited" = "orange","Retained" = "deepskyblue") )+ 
  scale_x_discrete(limits=c("Non-Travel","Travel_Rarely","Travel_Frequently"))+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))

```

There is also an increasing trend of attrition as the volume of business travel also increases.




<font size="6">  Classifying Attrition</font>

Naïve Bayes model was used in order to classify attrition.  75% of the employee data was split into a training set and the remaining was used as a test set in order to test the accuracy of the model.  The metrics used for the model were:Age, Business Travel, Department, Distance from Home, Education Level, Environment Satisfaction, Hourly Rate, Job Involvement, Job Level, Job Role, Job Satisfaction, Marital Status, Monthly Income, Overtime, Performance Rating, Relationship Satisfaction, Total Working Years, Work Life Balance, Years At Company, Years In Current Role, Years Since Last Promotion, Years With Current Manager

In using these variables, we were able to get 88% accuracy in the prediction, sensitivity of 93.3% (the ability to properly classify people who attrited) and a specificity of 62% (the ability to properly classify people who did not attrite.

```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Naive Bayes

set.seed(1234)
splitPerc = .75

  trainIndices = sample(1:dim(model_data)[1],round(splitPerc * dim(model_data)[1]))
  train_attr_nb = model_data[trainIndices,]
  test_attr_nb = model_data[-trainIndices,]

nb_model <- naiveBayes(train_attr_nb[,c(1,3,5,6,7,9,11,12,13,14,15,16,17,20,22,23,25,27,28,29,30,31)], train_attr_nb$attrition)
t<-table(predict(nb_model, test_attr_nb[,c(1,3,5,6,7,9,11,12,13,14,15,16,17,20,22,23,25,27,28,29,30,31)]),test_attr_nb$attrition)  
CM = confusionMatrix(t)
CM

nb_model_final <-naiveBayes(model_data[,c(1,3,5,6,7,9,11,12,13,14,15,16,17,20,22,23,25,27,28,29,30,31)], model_data$attrition)
attrition_predict<-predict(nb_model_final, CS2_attrition_test[,c(2,3,5,6,7,11,13,14,15,16,17,18,19,23,25,26,29,31,32,33,34,35)])
attrition_predict_nb <- data.frame(CS2_attrition_test$id,attrition_predict)
names(attrition_predict_nb) <-c('ID','Attrition')
write.csv(attrition_predict_nb,'C:/Users/rosjo/Documents/Masters Program/Doing Data Science/Assignments/Case Study 2/Attrition_Classification.csv')
```


<font size="6">  Additional Insights</font>

```{r,message = FALSE,warning = FALSE, echo=FALSE}
## Additional Insights

model_datasalarysummary = model_data%>% 
  mutate(monthlyincomerange = case_when(
    between (monthlyincome,0,3999.99)~"Less than $4K",
    between (monthlyincome,4000,8000)~"Between $4K and $8K",
    between (monthlyincome,8000.01,12000)~"Between $8K and $12K",
    between (monthlyincome,12000.01,16000)~"Between $12K and $16K",
    between (monthlyincome,16000.01,20000)~"Between $16K and $20K",
    monthlyincome > 20000 ~"Greater than $20K"
  ))
monthlyincomerange<-as.factor(model_datasalarysummary$monthlyincomerange)

##Plotting Job Role vs. Job Satisfaction
cross_jrjs <- table(jobsatisfaction, jobrole)
plot_data_jrjs<- as.data.frame(round(prop.table(cross_jrjs,2)*100, digits = 0))

plot_data_jrjs%>%
  ggplot(aes(x=jobrole, y=Freq, fill = jobsatisfaction))+ geom_bar(stat="identity")+
  geom_text(aes(label  = paste0(Freq,"%")), position=position_stack(vjust=0.5), size = 3.2, color = "black")+
  ggtitle('Percentage Job Satisfaction by Job Role')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Not Satisfied" = "orange","Somewhat Satisfied" = "khaki2", "Satisfied" = "deepskyblue", "Very Satisfied" = "hotpink2") )+
  scale_x_discrete(labels = function(jobrole) str_wrap(jobrole, width = 10))+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust =1, size = 8))
```

Generally all roles are more than 50% Satisfied or Very Satisfied (3 or 4 in response).  Manager roles were at 49% in those two categories.



```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Job Level vs. Job Satisfaction
cross_jljs <- table(jobsatisfaction, joblevel)
plot_data_jljs<- as.data.frame(round(prop.table(cross_jljs,2)*100, digits = 0))

plot_data_jljs%>%
  ggplot(aes(x=joblevel, y=Freq, fill = jobsatisfaction))+ geom_bar(stat="identity")+
  geom_text(aes(label  = paste0(Freq,"%")), position=position_stack(vjust=0.5), size = 3.2, color = "black")+
  ggtitle('Percentage Job Satisfaction by Job Level')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Not Satisfied" = "orange","Somewhat Satisfied" = "khaki2", "Satisfied" = "deepskyblue", "Very Satisfied" = "hotpink2") )+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(size = 10))
```

We do see a decreasing trend of job satisfaction as the job level’s increase with Level 5 roles having the least proportion of employees rating a 4 in Job Satisfaction.



```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Plotting Monthly Salary vs. Job Satisfaction
cross_mirjs <- table(jobsatisfaction, monthlyincomerange)
plot_data_mirjs<- as.data.frame(round(prop.table(cross_mirjs,2)*100, digits = 0))

plot_data_mirjs%>%
  ggplot(aes(x=monthlyincomerange, y=Freq, fill = jobsatisfaction))+ geom_bar(stat="identity")+
  geom_text(aes(label  = paste0(Freq,"%")), position=position_stack(vjust=0.5), size = 3.2, color = "black")+
  ggtitle('Percentage Job Satisfaction by Monthly Income Range')+
  ylab('Percentage')+
  scale_fill_manual("attrition",values = c("Not Satisfied" = "orange","Somewhat Satisfied" = "khaki2", "Satisfied" = "deepskyblue", "Very Satisfied" = "hotpink2") )+
   scale_x_discrete(limits=c("Less than $4K","Between $4K and $8K","Between $8K and $12K","Between $12K and $16K","Between $16K and $20K"), labels = function(monthlyincomerange) str_wrap(monthlyincomerange, width = 10))+
  theme(legend.position = "bottom",
        legend.text = element_text(size=10),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y =  element_blank(),
        title = element_text(face="bold", color = "black", size = 8),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust =1, size = 9))

```

Similar to job levels, the higher paid employees then to be less satisfied.



<font size="6">  Predicting Salary</font>

A linear regression model was used in order to predict attrition.  Similar to classifying attrition, 75% of the employee data was split into a training set and the remaining was used as a test set in order to test the accuracy of the model.
The initial linear model was run without any filters on variables to determine the best variables to include.  Based on the results, the metrics used for the model were: Business Travel Frequency, Job Level, Job Role, Total Working Years and Years Since Last Promotion.

In using these variables, we were able to get a Root Mean Square Error (RMSE) of 1,122.  The RMSE measures how concentrated the data is around the line of best fit.


```{r,message = FALSE,warning = FALSE, echo=FALSE}

set.seed(1234)
splitPerc = .75

  trainIndices = sample(1:dim(model_data)[1],round(splitPerc * dim(model_data)[1]))
  train_salary_lm = model_data[trainIndices,]
  test_salary_lm = model_data[-trainIndices,]

model_salary <- lm(monthlyincome~.,data = train_salary_lm)
summary(model_salary)

model_salary_2 <- lm(monthlyincome~businesstravel + joblevel + jobrole + totalworkingyears + yearssincelastpromotion,data = train_salary_lm)
summary(model_salary_2)

pred_2<-predict.lm(model_salary_2,test_salary_lm)
salary_actual<-test_salary_lm$monthlyincome
salary_predicted<-pred_2
RMSE = rmse(salary_actual, salary_predicted)
RMSE

model_salary_final <- lm(monthlyincome~businesstravel + joblevel + jobrole + totalworkingyears + yearssincelastpromotion,data = model_data)
summary(model_salary_final)
pred_final<-predict.lm(model_salary_final,CS2_salary_test)

Salary_Prediction_Regress <- data.frame(CS2_salary_test$ï..id, pred_final)
names(Salary_Prediction_Regress) <-c('ID','MonthlyIncome')
write.csv(Salary_Prediction_Regress,'C:/Users/rosjo/Documents/Masters Program/Doing Data Science/Assignments/Case Study 2/Salary_Prediction.csv')
```

```{r,message = FALSE,warning = FALSE, echo=FALSE}

##Link to YouTube File: 

https://youtu.be/p-00V_oAiiw



```


